Mathematical appendix to "Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa

### 2023-03-01 src/docs_paper/appendix.Rmd
Journal formatting requirements

<!-- So that the tables and figures in this appendix are prefixed by "A." -->
\setcounter{table}{0}  
\renewcommand{\thetable}{A.\arabic{table}} 
\setcounter{figure}{0} 
\renewcommand{\thefigure}{A.\arabic{figure}}

@@

Let $i \in \{1, \ldots, n\}$ denote subnational units (Table B.5) which partition the 13 studied AGYW priority countries $c[i] \in \{1, \ldots, 13\}$.
We considered the years 1999-2018 denoted as $t \in \{1, \ldots, T\}$, and age groups $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$.
Let the four risk groups be $k \in \{1, 2, 3, 4\}$ and denote being in either the third or fourth risk group by $k = 3^{+}$.

@@

We performed model selection on the basis of the conditional predictive ordinate (CPO) criterion [@pettit1990conditional], selecting model M2.
This model included Besag spatial random effects and IID temporal random effects.
For reference, we also computed the deviance information criterion (DIC) [@spiegelhalter2002bayesian] and widely applicable information criterion (WAIC) [@watanabe2013widely] (Table \ref{tab:multinomial-model-performance} and Figure \ref{fig:multinomial-model-performance}).

@@

We selected the best model according to the CPO statistic, which was model L6.
CPO values, along with DIC and WAIC values for reference, are presented in Table \ref{tab:logistic-model-performance} and Figure \ref{fig:logistic-model-performance}.
Inclusion of Besag spatial random effects, rather than IID, consistently improved performance.
Benefits from inclusion of covariates were more marginal.
That said, as some countries had no suitable surveys, we preferred to include covariate information such that the estimates in these countries are based on some country-specific data.

@@

Figure \ref{fig:coverage} shows the PIT histogram and ECDF difference plots for our final model, faceted by risk group.
<!-- Comment on the results. -->
<!-- What is going on with the cohabiting partner group towards one? -->

@@

Next, we used the assumed $\text{Gamma}(\alpha = 10.4, \beta = 0.36)$ FSW age distribution in South Africa from the Thembisa model [@johnson2020thembisa] to calculate the implied ratio between the number of FSW and the sexually debuted population in each age group.
We assumed these ratios in South Africa were applicable to every country to calculate the number of FSW by age group in all 13 countries (Figure \ref{fig:age-disagg-fsw-line}).

@@

Given the fitted log odds ratios $\log(\text{OR}_k)$, we disaggregated Naomi estimates of PLHIV $H_{ia}$ on the logit scale using numerically optimisation to obtain the value of $\theta$ minimising the function
\begin{equation}
\hat \theta = \argmin_{\theta \in [-10, 10]} \left( \sum_{k} \left( \text{logistic}(\theta + \log(\text{OR}_k)) \cdot N_{iak} \right) - H_{ia} \right)^2
\end{equation}
where $\text{logistic}(x) = \exp(x) / (1 + \exp(x))$ such that $\text{logistic}(\hat \theta + \log(\text{OR}_k)) = \rho_{iak}$.
Figures B.18 - B.30 show the resulting HIV prevalence estimates by risk group in each country.
The number of PLHIV are then obtained by $H_{iak} = \rho_{iak} N_{iak}$, where $N_{iak}$ is the risk group population size.

@@

We calculated HIV incidence $\lambda_{iak}$ and number of new HIV infections $I_{iak}$ stratified according to district, age group and risk group by linear disaggregation
\begin{align}
    I_{ia} &= \sum_k I_{iak} = \sum_k \lambda_{iak}N_{iak} \\
    &= 0 + \lambda_{ia2} N_{ia2} + \lambda_{ia3} N_{ia3} + \lambda_{ia4} N_{ia4} \\
    &= \lambda_{ia2} \left(N_{ia2}  + \text{RR}_{3} N_{ia3} + \text{RR}_4(\lambda_{ia}) N_{ia4}  \right).
\end{align}
Risk group specific HIV incidence estimates are then given by
\begin{align}
    \lambda_{ia1} &= 0, \\
    \lambda_{ia2} &= I_{ia} / \left(N_{ia2} + \text{RR}_{3} N_{ia3} + \text{RR}_4(\lambda_{ia}) N_{ia4}\right), \\
    \lambda_{ia3} &= \text{RR}_{3} \lambda_{ia2}, \\
    \lambda_{ia4} &= \text{RR}_4(\lambda_{ia}) \lambda_{ia2}.
\end{align}
which we evaluated using Naomi model estimates of the number of new HIV infections $I_{ia} = \lambda_{ia} N_{ia}$, HIV infection risk ratios $\{\text{RR}_3, \text{RR}_4(\lambda_{ia})\}$, and risk group population sizes as above.
The risk ratio $\text{RR}_4(\lambda_{ia})$ was defined as a function of general population incidence.
Figures B.31 - B.43 show the resulting HIV incidence estimates by risk group in each country.
The number of new HIV infections are then $I_{iak} = \lambda_{iak} N_{iak}$.

@@

We calculated the number of new infections that would be reached prioritising according to each possible stratification of the population--that is for all $2^3 = 8$ possible combinations of stratification by location, age, and risk group. 
As an illustration, for stratification just by age, we aggregated the number of new HIV infections and HIV incidence as such 
\begin{align}
    I_a &= \sum_{ik} I_{iak}, \\
    \lambda_a &= I_a / \sum_{ik} N_{iak}.
\end{align}
Under this stratification, individuals in each age group $a$ are prioritised according to the highest HIV incidence $\lambda_a$.
By cumulatively summing the expected infections, for each fraction of the total population reached we calculated the fraction of total expected new infections that would be reached.
Figures B.44 - B.56 show the percentage of new infections that would be reached prioritising according to each possible stratification within each country.

>>>>

<!-- So that the tables and figures in this appendix are numbered alphabetically --->
\setcounter{table}{0}  
\renewcommand{\thetable}{\Alph{table}} 
\setcounter{figure}{0} 
\renewcommand{\thefigure}{\Alph{figure}}

@@

Let $i \in \{1, \ldots, n\}$ denote subnational units (Table E in S2 Text) which partition the 13 studied AGYW priority countries $c[i] \in \{1, \ldots, 13\}$.
We considered the years 1999-2018 denoted as $t \in \{1, \ldots, T\}$, and age groups $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$.
Let the four risk groups be $k \in \{1, 2, 3, 4\}$ and denote being in either the third or fourth risk group by $k = 3^{+}$.

@@

We performed model selection on the basis of the conditional predictive ordinate (CPO) criterion [@pettit1990conditional], selecting model M2.
This model included Besag spatial random effects and IID temporal random effects.
For reference, we also computed the deviance information criterion (DIC) [@spiegelhalter2002bayesian] and widely applicable information criterion (WAIC) [@watanabe2013widely] (Table \ref{tab:multinomial-model-performance} and Fig \ref{fig:multinomial-model-performance}).

@@

We selected the best model according to the CPO statistic, which was model L6.
CPO values, along with DIC and WAIC values for reference, are presented in Table \ref{tab:logistic-model-performance} and Fig \ref{fig:logistic-model-performance}.
Inclusion of Besag spatial random effects, rather than IID, consistently improved performance.
Benefits from inclusion of covariates were more marginal.
That said, as some countries had no suitable surveys, we preferred to include covariate information such that the estimates in these countries are based on some country-specific data.

@@

Fig \ref{fig:coverage} shows the PIT histogram and ECDF difference plots for our final model, faceted by risk group.
<!-- Comment on the results. -->
<!-- What is going on with the cohabiting partner group towards one? -->

@@

Next, we used the assumed $\text{Gamma}(\alpha = 10.4, \beta = 0.36)$ FSW age distribution in South Africa from the Thembisa model [@johnson2020thembisa] to calculate the implied ratio between the number of FSW and the sexually debuted population in each age group.
We assumed these ratios in South Africa were applicable to every country to calculate the number of FSW by age group in all 13 countries (Fig \ref{fig:age-disagg-fsw-line}).

@@

Given the fitted log odds ratios $\log(\text{OR}_k)$, we disaggregated Naomi estimates of PLHIV $H_{ia}$ on the logit scale using numerically optimisation to obtain the value of $\theta$ minimising the function
\begin{equation}
\hat \theta = \argmin_{\theta \in [-10, 10]} \left( \sum_{k} \left( \text{logistic}(\theta + \log(\text{OR}_k)) \cdot N_{iak} \right) - H_{ia} \right)^2
\end{equation}
where $\text{logistic}(x) = \exp(x) / (1 + \exp(x))$ such that $\text{logistic}(\hat \theta + \log(\text{OR}_k)) = \rho_{iak}$.
Fig R - AD in S2 Text show the resulting HIV prevalence estimates by risk group in each country.
The number of PLHIV are then obtained by $H_{iak} = \rho_{iak} N_{iak}$, where $N_{iak}$ is the risk group population size.

@@

We calculated HIV incidence $\lambda_{iak}$ and number of new HIV infections $I_{iak}$ stratified according to district, age group and risk group by linear disaggregation
\begin{align}
    I_{ia} &= \sum_k I_{iak} = \sum_k \lambda_{iak}N_{iak} \\
    &= 0 + \lambda_{ia2} N_{ia2} + \lambda_{ia3} N_{ia3} + \lambda_{ia4} N_{ia4} \\
    &= \lambda_{ia2} \left(N_{ia2}  + \text{RR}_{3} N_{ia3} + \text{RR}_4(\lambda_{ia}) N_{ia4}  \right).
\end{align}
Risk group specific HIV incidence estimates are then given by
\begin{align}
    \lambda_{ia1} &= 0, \\
    \lambda_{ia2} &= I_{ia} / \left(N_{ia2} + \text{RR}_{3} N_{ia3} + \text{RR}_4(\lambda_{ia}) N_{ia4}\right), \\
    \lambda_{ia3} &= \text{RR}_{3} \lambda_{ia2}, \\
    \lambda_{ia4} &= \text{RR}_4(\lambda_{ia}) \lambda_{ia2}.
\end{align}
which we evaluated using Naomi model estimates of the number of new HIV infections $I_{ia} = \lambda_{ia} N_{ia}$, HIV infection risk ratios $\{\text{RR}_3, \text{RR}_4(\lambda_{ia})\}$, and risk group population sizes as above.
The risk ratio $\text{RR}_4(\lambda_{ia})$ was defined as a function of general population incidence.
Fig AE - AQ in S2 Text show the resulting HIV incidence estimates by risk group in each country.
The number of new HIV infections are then $I_{iak} = \lambda_{iak} N_{iak}$.

@@

We calculated the number of new infections that would be reached prioritising according to each possible stratification of the population--that is for all $2^3 = 8$ possible combinations of stratification by location, age, and risk group. 
As an illustration, for stratification just by age, we aggregated the number of new HIV infections and HIV incidence as such 
\begin{align}
    I_a &= \sum_{ik} I_{iak}, \\
    \lambda_a &= I_a / \sum_{ik} N_{iak}.
\end{align}
Under this stratification, individuals in each age group $a$ are prioritised according to the highest HIV incidence $\lambda_a$.
By cumulatively summing the expected infections, for each fraction of the total population reached we calculated the fraction of total expected new infections that would be reached.
Fig AR - BD in S2 Text show the percentage of new infections that would be reached prioritising according to each possible stratification within each country.
